A comprehensive framework to capture the arcana of neuroimaging analysis

### 2019-03-20 manuscript/manuscript.tex
final touch ups before sending to co-authors

Entries in the data specification can be of either a
\emph{file-set} or \emph{field} type. File-sets represent single files,
or sets of related files typically contained within a directory
(e.g. a multi-volume DICOM dataset). Fields
represent integer, floating point, date-time, or unicode string
values. By default, a field contains a single value but if the \emph{array}
flag is set, a field can contain a list of values.
Each file set references a \emph{FileFormat} object, which
specifies the formats of the files in the set. File formats are
explicitly registered by the researcher at design time using the
\emph{FileFormat.register(format)} class method to avoid conflicts where
the same extension is used for different formats in different contexts.

@@

\begin{figure}
\lstinputlisting[basicstyle=\footnotesize,firstline=2, lastline=59]{../code/example_study.py}
\caption{Example data and parameter specifications. The data
specification specifies two input file sets, `one' and `ten' and
ten derived file sets that can be derived from them, at least
indirectly. Each derived data spec, specifies the name of the pipeline
constructor that creates the pipeline that derives them. Parameter
specifications specify a name and default value for free parameters of
the Study class}
\label{fig:data_spec}
\end{figure}

@@

While derivatives generated by a Study object are named in accordance
with the data specification of the Study class, input file sets and fields
do not need specific names when using basic or XNAT
repositories. In these cases, a selection stage matches input data to
entries in the data specification of the Study class. The criteria for
these selections are passed to the
the \emph{inputs} argument of the Study object at instantiation in a
dictionary mapping data specification names to \emph{FilesetInput} and
\emph{FieldInput} objects.

@@

Specific file sets and fields can be passed as inputs to a Study in
\emph{FilesetCollection} and \emph{FieldCollection}
objects. Collection objects reference an entry in the data specification
and contain a single Fileset or Field for every session (or
subject/visit/study depending on the frequency of the corresponding data
specifcation). Collection objects can be used to pass reference atlases and
templates as inputs to analyses. They can also be set as the default
for a data specification entry via the \emph{default} keyword
argument. However, for the sake of portability, default inputs should be
restricted to data in publically accessible repositories or those included in
standard software packages (e.g. FSL).

@@

We present Arcana, a software framework to facilitate the development of
analysis suites for neuroimaging data that implement
complete workflows from repository data to publication results. The
encapsulation of repository data and workflow generation in Arcana
enables researchers to create robust workflows while focussing on the
core logic of their analysis. Arcana's modular pipeline and OO
architecture promotes code reuse between different workflows by
facilitating the sharing of common segments (e.g. registration,
segmentation). The clear separation of analysis design from its
application leads to portable workflows, which can be applied to
datasets stored in a number of storage systems. In addition, the
management of intermediate derivatives, provenance and software
versioning, coupled with ability to submit jobs to HPC clusters, enables
workflows implemented in Arcana to scale to large datasets. Researchers
are able to quickly prototype analyses that can be deployed on large-scale
infrastructure on local workstations without modification.

@@

The framework provided by Arcana should facilitate the development of workflows
that are consistent with the F.A.I.R. guiding principles \citep{wilkinson_fair_2016}.
Each pipeline is accompanied by a text description and
a list of citations to describe the techniques it employs.
The extensibility and modularity of workflows in Arcana should encourage
their reuse, and enable communal development of code repositories.
Arcana analyses are interoperable between different
repository and execution environments. In addition, the development of
Banana is planned to closely follow the specification for BIDS
derivatives \citep{gorgolewski_bids_nodate} to enable interoperability with
other BIDS apps.

>>>>

Entries in the data specification can be of either a
\emph{file-set} or \emph{field} type. File-sets represent single files,
or sets of related files typically contained within a directory
(e.g. a multi-volume DICOM dataset). Fields
represent integer, floating point, date-time, or unicode string
values. By default, a field contains a single value but if the \emph{array}
flag is set, a field can contain a list of values.
Each file-set references a \emph{FileFormat} object, which
specifies the formats of the files in the set. File formats are
explicitly registered by the researcher at design time using the
\emph{FileFormat.register(format)} class method to avoid conflicts where
the same extension is used for different formats in different contexts.

@@

\begin{figure}
\lstinputlisting[basicstyle=\footnotesize,firstline=2, lastline=59]{../code/example_study.py}
\caption{Example data and parameter specifications. The data
specification specifies two input file-sets, `one' and `ten' and
ten derived file-sets that can be derived from them, at least
indirectly. Each derived data spec, specifies the name of the pipeline
constructor that creates the pipeline that derives them. Parameter
specifications specify a name and default value for free parameters of
the Study class}
\label{fig:data_spec}
\end{figure}

@@

While derivatives generated by a Study object are named in accordance
with the data specification of the Study class, input file-sets and fields
do not need specific names when using basic or XNAT
repositories. In these cases, a selection stage matches input data to
entries in the data specification of the Study class. The criteria for
these selections are passed to the
the \emph{inputs} argument of the Study object at instantiation in a
dictionary mapping data specification names to \emph{FilesetInput} and
\emph{FieldInput} objects.

@@

Specific file-sets and fields can be passed as inputs to a Study in
\emph{FilesetCollection} and \emph{FieldCollection}
objects. Collection objects reference an entry in the data specification
and contain a single Fileset or Field for every session (or
subject/visit/study depending on the frequency of the corresponding data
specifcation). Collection objects can be used to pass reference atlases and
templates as inputs to analyses. They can also be set as the default
for a data specification entry via the \emph{default} keyword
argument. However, for the sake of portability, default inputs should be
restricted to data in publically accessible repositories or those included in
standard software packages (e.g. FSL).

@@

We present Arcana, a software framework to facilitate the development of
analysis suites for neuroimaging data that implement
complete workflows from repository data to publication results. The
encapsulation of repository data and workflow generation in Arcana
enables researchers to create robust workflows while focussing on the
core logic of their analysis. Arcana's modular pipeline and OO
architecture promotes code reuse between different workflows by
facilitating the sharing of common segments (e.g. registration,
segmentation). The clear separation of analysis design from its
application leads to portable workflows, which can be applied to
datasets stored in a number of storage systems. In addition, the
management of intermediate derivatives, provenance and software
versioning, coupled with ability to submit jobs to HPC clusters, enables
workflows implemented in Arcana to scale to large datasets. Researchers
are able to quickly prototype analyses that can be deployed to large-scale
infrastructure on local workstations without modification.

@@

The framework provided by Arcana should facilitate the development of workflows
that are consistent with the F.A.I.R. guiding principles \citep{wilkinson_fair_2016}.
Each pipeline is accompanied by a text description and
a list of citations to describe the techniques it employs.
The extensibility and modularity of workflows in Arcana should encourage
their reuse, and enable communal development of code repositories.
Arcana analyses are interoperable between different
repository and execution environments. In addition, the development of
Banana is planned to closely follow the specification for BIDS
derivatives \citep{gorgolewski_bids_nodate} for interoperability with
other BIDS apps.
